Graph Product Structure for Non-Minor-Closed Classes

### 2021-11-19 slides/slides.tex
Pffft

\begin{frame}
    \frametitle{Abstraction: Shortcut Systems}

    \begin{itemize}
        \item \emph{$(k,d)$-shortcut system} for $G$:
        \begin{itemize}
            \item<2-> a set $\mathcal{P}$ of length-$\le\!\! k$ paths in $G$
            \item<2-> each $v\in V(G)$ appears interior to at most $d$ paths in $\mathcal{P}$
        \end{itemize}
        \begin{center}
            \multiinclude[<+>][format=pdf,start=1]{figs/shortcut-system}%
        \end{center}
        \item<3-> $G^{\mathcal{P}}:= G\cup\{vw:\mbox{$v$ and $w$ are endpoints of a path in $\mathcal{P}$}\}$
    \end{itemize}
\end{frame}

@@

    If $G'$ is $k$-planar then $G'\subseteq G^{\mathcal{P}}$ where
    \begin{itemize}
        \item $G$ is planar
        \item $\mathcal{P}$ is a $(k+1,2)$-shortcut system
    \end{itemize}
    \begin{center}
        \multiinclude[<+>][format=pdf,start=1]{figs/kgp}%
    \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Proof: The Picture}
  \begin{itemize}
    \item $G= H\boxtimes P\boxtimes K_{\ell}$
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Shortcuts over multiple layers}

  \begin{itemize}
    \item Shortcuts in $\mathcal{P}$ may jump $k$ layers
    \item Easy solution: Compress $k$ layers into one
    \begin{itemize}
      \item Length $k-1$ path $P'$ in $P$ compresses down to vertex
      \item $P'\boxtimes K_{\ell}\subseteq K_{\ell k}$
    \end{itemize}
    \item Increases $\ell$ by a factor of $k$
    \item $\mathcal{P}$ is a $(k,d)$-shortcut system over $H\boxtimes P\boxtimes K_{\ell k}$ with all shortcuts crossing $\le\!\! 1$ layers
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item Contract each copy of $K_{k\ell}$, so we are left with $H\boxtimes P$
    \item Each shortcut in $\mathcal{P}$ projects onto a walk in $H\boxtimes P$
    \item Each vertex of $H\boxtimes P$ is used in at most $dk\ell$ shortcuts
    \item Studying a $(k,dk\ell)$-shortcut system $\mathcal{P'}$ over $G'=H\boxtimes P$
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Nice Tree Decompositions}

  \begin{itemize}
    \item A \emph{nice} rooted tree decomposition $\mathcal{T}=(B_x:x\in V(T))$ of $H$ has 
    \begin{itemize}
      \item $V(T) = V(H)$
      \item $x$ is lowest common ancestor of $\{x'\in V(T):x\in B_{x'}\}$
    \end{itemize}
    \begin{center}
      \includegraphics{figs/nice-td}
    \end{center}
    \item If $xy\in E(H)$ then $x\prec_T y$ or $y\prec_T x$
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item<2-> Start with $B_x\gets\{x\}$ for each $x\in V(T)$
    \item<3-> If $xy\in E(H)$ and $x\prec_T y$ then drag $x$ down into $B_y$
  \end{itemize}
  \begin{center}
    \multiinclude[<+>][format=pdf,start=1]{figs/alternative-view}%
  \end{center}
\end{frame}

@@

  \textbf{Lemma (Pilipczuk--Siebertz):}\footnote{Yes, the same paper!} For any $z\in V(H)$ the number of $x\in V(H)$ such that $x\prec_T z\prec_T z'$ and $\mathop{dist}_H(x,z')\le k$ is at most $\binom{k+t}{t}$.
  \begin{center}
    \includegraphics{figs/pz}%
  \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Lifting in $T$}
  \begin{itemize}
    \item For $v$ in $G$, $a(v)$ is the LCA of $\cup\{S_I:I\in\mathcal{P'},\, v\in V(I)\}$
    \item For $x\in V(T)$, $S_x:=\{v\in V(G): a(v)=x\}$
    \begin{center}
      \includegraphics{figs/nice-td-6}%
    \end{center}
    \item Note: $\mathcal{S}:=(S_x:x\in V(T))$ is a partition of $V(G)$
\end{itemize}
\end{frame}

@@

    Let $H':=G/\mathcal{S}$

    \textbf{Claim:} $\tw(H')\le\binom{k+t}{t}-1$
    \vspace{1cm}
    \textit{Proof:}
    \begin{itemize}
      \item Each $x\in V(H')$ is obtained by contracting $S_x$
      \item Use tree nice tree decomposition with the same tree $T$
      \item If $xz\in E(H')$ and $x\prec_T z$ then drag $x$ down into $B_z$
      \item Point: If $x\in B_{z'}$ then $x\prec_T z'\prec_T z$ and $\mathop{dist}_{H}(x,z)\le k$.  
      \item By Pilipczuk-Siebertz Lemma, $|B_{z'}|\le \binom{k+t}{t}$.
    \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Thickness}

  \begin{itemize}
    \item Recall that $\mathcal{P'}$ is a $(k,dk\ell)$-shortcut system over $H\boxtimes P$.
    \item We can show that the subset of $H\boxtimes P\boxtimes K_{\ell}$ corresponding to $S_x$ contains at most $d\ell(k^2+3)$ vertices from each layer of $H\boxtimes P\boxtimes K_{\ell}$.
    \item Implies $G^{\mathcal{P}}\subseteq H'\boxtimes P\boxtimes K_{d\ell(k^2+3)}$
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
    \frametitle{$(k,d)$-Shortcut Systems}
    \framesubtitle{An abstraction of $k$-planarity}

    \begin{itemize}
        \item \emph{$(k,d)$-shortcut system} for $G$:
        \begin{itemize}
            \item<2-> a set $\mathcal{P}$ of length-$\le\!\! k$ paths in $G$
            \item<2-> each $v\in V(G)$ appears interior to at most $d$ paths in $\mathcal{P}$
        \end{itemize}
    \end{itemize}
    \begin{center}
        \multiinclude[<+>][format=pdf,start=1]{figs/shortcut-system}%
        \uncover<2->{\\ $(3,2)$-shortcut system}
    \end{center}
    \begin{itemize}
        \item<3-> $G^{\mathcal{P}}:= G\cup\{vw:\mbox{$v$ and $w$ are endpoints of a path in $\mathcal{P}$}\}$
    \end{itemize}
\end{frame}

@@

    If $G$ is $k$-planar then $G\subseteq G_0^{\mathcal{P}}$ where
    \begin{itemize}
        \item<2-> $G_0$ is planar
        \item<3-> $\mathcal{P}$ is a $(k+1,2)$-shortcut system
    \end{itemize}
    \begin{center}
        \multiinclude[<+>][format=pdf,start=1]{figs/kgp}%
    \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Mental Map of $H\boxtimes P\boxtimes K_{\ell}$}
  \begin{itemize}
    \item $H\uncover<2->{\boxtimes P\uncover<3->{\boxtimes K_{\ell}}}$
  \end{itemize}
  \begin{center}
    \multiinclude[<+>][format=pdf,start=1]{figs/product-x}%
  \end{center}
\end{frame}

\begin{frame}
  \frametitle{Shortcuts over multiple rows}

  \begin{itemize}
    \item Shortcuts in $\mathcal{P}$ may jump $k$ rows
    \item<2-> Easy solution: Compress $k$ rows into one
  \end{itemize}
    \begin{center}
    \only<1>{\includegraphics{figs/layer-compression-1}}%
    \only<2->{\includegraphics{figs/layer-compression-2}}%
    \end{center}
  \begin{itemize}
    \item<3-> Increases $\ell$ by a factor of $k$
    \item<4-> $\mathcal{P}$ is a $(k,d)$-shortcut system over $H\boxtimes P\boxtimes K_{\ell k}$ with all shortcuts crossing $\le\!\! 1$ layers
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item<2-> Contract each copy of $K_{k\ell}$, so we are left with $H\boxtimes P$
  \end{itemize}
    \begin{center}
    \only<1>{\includegraphics{figs/layer-compression-2}}%
    \only<2->{\includegraphics{figs/layer-compression-3}}%
    \end{center}
  \begin{itemize}
    \item<3-> Each vertex of $H\boxtimes P$ is used in at most $dk\ell$ shortcuts
    \item<4-> $(k,dk\ell)$-shortcut system $\mathcal{P'}$ over $G'=H\boxtimes P$
    \item<4-> shortcuts span $\le\!\!1$ rows
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Nice Tree Decompositions}
  \framesubtitle{Nodes of $T$ are same as vertices of $H$}

  \begin{itemize}
    \item A \emph{nice} rooted tree decomposition $\mathcal{T}=(B_x:x\in V(T))$ of $H$ has
    \begin{itemize}
      \item $V(T) = V(H)$
      \item $x$ is lowest common ancestor of $\{x'\in V(T):x\in B_{x'}\}$
    \end{itemize}
    \begin{center}
      \includegraphics[scale=.8]{figs/nice-td}
    \end{center}
    \item If $xy\in E(H)$ then $x\prec_T y$ or $y\prec_T x$
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
    \item<1-> Start with $T$ such that $x\mathbin{\diamond}_T y$ for each $xy\in E(H)$
    \item<2-> $B_x\gets\{x\}$ for each $x\in V(T)$
    \item<3-> If $xy\in E(H)$ and $x\prec_T y$ then drag $x$ down into $B_y$
  \end{itemize}
  \begin{center}
    \multiinclude[<+>][format=pdf,start=1]{figs/alternative-view}%
  \end{center}
\end{frame}

@@

  \textbf{Lemma (Pilipczuk--Siebertz):}\footnote{Yes, the same paper!} For any $z\in V(H)$ the number of $x\in V(H)$ such that $x\prec_T z\prec_T z'$ and $\mathop{dist}_H(x,z')\le k$ is at most $\binom{k+t}{t}$.
  \begin{center}
    \includegraphics[scale=.8]{figs/pz}%
  \end{center}
  \begin{itemize}
      \item<2->Problem: $vw$-shortcut in $H\boxtimes P$ projects onto $xz'$ shortcut in $H$, but $x\not\preceq_T z'$ and $z'\not\preceq_T x$
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{Lifting in $T$}
  \begin{itemize}
    \item For $v$ in $H\boxtimes P$, $a(v)$ is the LCA of $\cup\{S_I:I\in\mathcal{P'},\, v\in V(I)\}$
    \item For $x\in V(T)$, $S_x:=\{v\in V(G): a(v)=x\}$
    \begin{center}
      \includegraphics[scale=.8]{figs/nice-td-6}%
    \end{center}
    \item<2-> $\mathcal{S}:=(S_x:x\in V(T))$ is a partition of $V(G)$
    \item<3->[$(\star)$] For any $vw$-shortcut in $H\boxtimes P$, $a(v)\preceq_T a(w)$ or $a(w)\preceq_T a(v)$
\end{itemize}
\end{frame}

@@

  \begin{itemize}
      \item[$(\star)$] For any $vw$-shortcut in $H\boxtimes P$, $a(v)\preceq_T a(w)$ or $a(w)\preceq_T a(v)$
  \end{itemize}

    \vspace{.5cm}
    Let $H':=G/\mathcal{S}$
    \vspace{.5cm}

    \textbf{Claim:} $\tw(H')\le\binom{k+t}{t}-1$\\
    \vspace{.5cm}
    \uncover<2->{\textit{Proof:}}
    \begin{itemize}
      \item<2-> Each $x\in V(H')$ is obtained by contracting $S_x$
      \item<3-> Start with $T$ from nice tree decomposition of $H$ (not $H'$)
      \item<4-> For each $xz\in E(H')$ where $x\prec_T z$, drag $x$ down into $B_z$
      \item<5-> Correctness follows from $(\star)$
      \item<6-> Point: If $x\in B_{z'}$ then
      \begin{itemize}
          \item $x\prec_T z'\prec_T z$ and
          \item $\mathop{dist}_{H}(x,z)\le k$ (because shortcuts have length $\le\!\!k$).
      \end{itemize}
      \item<7-> By Pilipczuk-Siebertz Lemma, $|B_{z'}|\le \binom{k+t}{t}$.
      \hfill{\qed}
    \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Thickness}
  \framesubtitle{Detail checking}

  \begin{itemize}
    \item Recall that $\mathcal{P'}$ is a $(k,dk\ell)$-shortcut system over $H\boxtimes P$.
    \item We can show that the subset of $H\boxtimes P\boxtimes K_{\ell}$ corresponding to $S_x$ contains at most $d\ell(k^2+3)$ vertices from each layer of $H\boxtimes P\boxtimes K_{\ell}$.
    \item Implies $G^{\mathcal{P}}\subseteq H'\boxtimes P\boxtimes K_{d\ell(k^3+3k)}$
  \end{itemize}
\end{frame}

\begin{frame}

\end{frame}

% \begin{frame}
%
%
% \end{frame}
%
